---
title: "Regression Models and Significance"
author: "Calvin Williamson, Science and Math"
subtitle: "Fall 26"
format:
  revealjs:
    theme: simple
    transition: slide
    slide-number: true
    css: RegressionModelsAndSignificance.css
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
#| include: false
relationships <- data.frame(
  slope = c(1, -1, 0.5, -0.5, 0.2, -0.2),
  intercept = c(0, 10, 2, 8, 4, 6),
  noise = c(1, 1, 1, 2, 2, 2)
)

make_datasets <- function(seed, n) {
  set.seed(seed)
  order <- sample(seq_len(nrow(relationships)))
  lapply(order, function(i) {
    x <- seq(0, 10, length.out = n)
    y <- relationships$slope[i] * x + relationships$intercept[i] +
      rnorm(n, 0, relationships$noise[i])
    if (n != 4) y <- pmin(pmax(y, 0), 10)
    data.frame(x = x, y = y)
  })
}

show_six <- function(datasets, title, point_size) {
  old_par <- par(mfrow = c(2, 3), mar = c(2.8, 2.8, 2, 0.5), oma = c(0, 0, 0, 0))
  on.exit(par(old_par))
  for (i in seq_along(datasets)) {
    d <- datasets[[i]]
    plot(d$x, d$y, main = paste("Dataset", i), xlab = "X", ylab = "Y",
      pch = 19, cex = point_size, col = "#2C7FB8")
    grid(col = "#DDDDDD")
  }
  if (nzchar(title)) mtext(title, outer = TRUE, cex = 1.35, font = 2)
}

model_results <- function(datasets) {
  rows <- lapply(seq_along(datasets), function(i) {
    fit <- lm(y ~ x, data = datasets[[i]])
    test <- summary(fit)$coefficients["x", ]
    data.frame(
      Dataset = i,
      Slope = unname(coef(fit)["x"]),
      Correlation = cor(datasets[[i]]$x, datasets[[i]]$y),
      `P-value` = unname(test["Pr(>|t|)"]),
      Significant = ifelse(test["Pr(>|t|)"] < .05, "Yes", "No"),
      check.names = FALSE
    )
  })
  results <- do.call(rbind, rows)
  rownames(results) <- NULL
  results
}

print_results <- function(results) {
  shown <- results[c("Dataset", "P-value")]
  decimal_p <- function(p) {
    places <- max(10, ceiling(-log10(p)) + 3)
    formatC(p, format = "f", digits = places)
  }
  shown$`P-value` <- vapply(shown$`P-value`, decimal_p, character(1))
  knitr::kable(shown, align = c("l", "r"))
}

datasets_75 <- make_datasets(42, 75)
datasets_10 <- make_datasets(34, 10)
datasets_4 <- make_datasets(12, 4)
results_75 <- model_results(datasets_75)
results_10 <- model_results(datasets_10)
results_4 <- model_results(datasets_4)
```

## Significant ($p < .05$)

- There is a **significant** linear relationship between $y$ and $x$.
- You **can use the regression equation** for predictions, provided the accuracy is good enough.
- Graphically, there are enough data to be convinced about the trendline.

We say: *The model is significant.*

## Not Significant ($p \geq .05$)

- There is **not a significant** linear relationship between $y$ and $x$.
- You **cannot use the regression equation** for predictions.
- There may be too little data or too much scatter to be convinced about the trendline.

We say: *The model is not significant.*

##

::: {.question-line}
**$n=75$** &nbsp;&nbsp; Which are significant? Smallest p-value? Largest p-value?
:::

```{r}
#| fig-width: 15
#| fig-height: 9.5
#| out-width: 100%
#| fig-align: center
show_six(datasets_75, "", 0.75)
```

## Answers: 75 Observations

```{r}
print_results(results_75)
```

##

::: {.question-line}
**$n=10$** &nbsp;&nbsp; Which are significant? Smallest p-value? Largest p-value?
:::

```{r}
#| fig-width: 15
#| fig-height: 9.5
#| out-width: 100%
#| fig-align: center
show_six(datasets_10, "", 1.25)
```

## Answers: 10 Observations

```{r}
print_results(results_10)
```

##

::: {.question-line}
**$n=4$** &nbsp;&nbsp; Which are significant? Smallest p-value? Largest p-value?
:::

```{r}
#| fig-width: 15
#| fig-height: 9.5
#| out-width: 100%
#| fig-align: center
show_six(datasets_4, "", 1.8)
```

## Answers: 4 Observations

```{r}
print_results(results_4)
```

With very small samples, even an apparent trend may not provide enough evidence.

## Sample Size and Significance

The same general strength of relationship can lead to different conclusions when the sample size changes.

- More observations give the test more information.
- Fewer observations make the location of the trendline less certain.
- The p-value combines the strength of the pattern with the amount of data.

::: {.notes}
Ported from RegressionModelsAndSignificance.qmd in calvinw/Ma322Notebooks. The answers use the standard two-sided t-test for the slope in simple linear regression.
:::
